Gelation as condensation frustrated by hydrodynamics and mechanical tension

### 2015-10-19 manuscript4.tex
Minor English fixes gravitational Peclet number

The mechanism of spinodal gelation has particularly been studied for colloidal suspensions since they can be regarded as a model system for other systems. Thanks to the extensive study in past, we now have a reasonable physical picture: Gelation is induced by spinodal decomposition to two phases whose dynamics are very different \cite{tanaka1999colloid,tanaka2000viscoelastic} and the resulting vitrification of the slower phase \cite{piazza1994phase,verhaegh1997,tanaka1999colloid,poon2002,lu2008gelation}. 
However, the knowledge of ordinary spinodal decomposition cannot explain the observed behaviour since it predicts that the minority colloid-rich phase should form isolated clusters rather than a percolated network. This mystery was solved by using an analogy of phase separation of colloidal suspensions to that of polymer solutions on the basis of the concept of dynamic asymmetry \cite{tanaka1999colloid}. 
What is common in these two systems is that a mixture is composed of large slow components (colloid or polymer) and small fast components (solvent). 
On a macroscopic level, such dynamically asymmetric mixtures are known to show universal pattern formation, known as ``viscoelastic phase separation'' \cite{tanaka2000viscoelastic}. 
On a microscopic level, however, there are essential differences between colloidal suspensions and polymer solutions coming from the topological difference (sphere vs. chain), leading to 
distinct rheological behaviours. Thus there are many open questions about the elementary process of phase separation of colloidal suspensions and the resulting colloidal gelation.   
From the physical viewpoint, the presence of a fluid component indicates the importance of fluid momentum conservation in phase separation, which makes hydrodynamics and mechanics active in structural evolution. This is the key for viscoelastic phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic}. There have been some numerical studies on the role of hydrodynamics~\cite{tanaka2000simulation,Furukawa2010} and mechanics~\cite{Tanaka2007,colombo2014self,testard2014intermittent} in colloidal gelation, however, there has so far been no experimental study that directly addresses these effects by dynamical observation of the process.

@@

To access roles of mechanics in the late stage, we realize the density matching of the order of 10$^{-4}$ and use a very accurate particle tracking method \cite{leocmach2013novel}. 

@@

The initial similarity between the trajectories of dilute gels and cluster phase is striking, as if their dynamics were ruled by the same phenomenon. To confront this hypothesis, we compute the time dependent static structure factor $S(q,t)$. For all gel and cluster samples, we observe the appearance of a low $q$ peak in $S(q)$, see Supplementary Figure XXX, which is characteristic of spinodal decomposition in a system of a conserved order parameter. To track the position of this peak, we compute the characteristic wave number $\langle q \rangle$ and show its temporal evolution in Fig.~\ref{fig:phasediag}c. The curves for all samples follow a master curve coherent with spinodal decomposition kinetics: At short times $\langle q \rangle(t)$ shows a plateau indicating that the low $q$ peak builds up at constant wave number corresponding to distances of about $2\sigma$. This plateau is characteristic of the early stage spinodal decomposition, which is described by Cahn's linear theory \cite{onuki2002phase}. At intermediate times, on the other hand, we observe coarsening with \textcolor{blue}{$\langle q \rangle \sim t^{-1/2}$, which is often observed in viscoelastic phase separation 
(see, e.g., \cite{Furukawa2010}).} Finally at longer times each sample deviates from the master curve to form a plateau indicating arrest. The more dilute samples arrest sooner, 
but the origin of arrest is different between clusters and percolated networks. For clusters, density fluctuations grow rather rapidly in the initial stage, but after the formation of isolated clusters it takes a long time since the diffusive evaporation-condensation mechanism does not work for colloidal viscoelastic phase separation and direct cluster collisions and coalescence are necessary for coarsening \cite{tanaka2000viscoelastic}. For networks, densification due to phase separation leads to dynamic arrest due to vitrification \cite{piazza1994phase,verhaegh1997,tanaka1999colloid,poon2002,zaccarelli2007,lu2008gelation} or formation of locally stable structures \cite{royall2008g}. 
Our observations indicate that both cluster and gel phases are due to viscoelastic spinodal decomposition: network-type spinodal for the gel, and droplet-type spinodal for the clusters where strong dynamical asymmetry between colloids and the solvent leads to unique roles of hydrodynamics and mechanics in phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic}.

@@

\clearpage
\begin{figure*}
	\includegraphics[width=16cm]{figure1.pdf}
	\caption{\textbf{Reservoir cell.} \textbf{a,} Sketch of our experimental setup. The observation cell contains initially colloids, polymer and no salt. \textbf{b,} Confocal slice of a gel formed \textit{in situ} by our method ($\phi=25.5~\%$, $c_p=1.4$ g/$\ell$), 1 hour after gelation. \textbf{c,} Idem for a gel at the same state point formed \textit{ex situ} and immediately pumped into a capillary. \textbf{d,} Gelation observed by our method. Experimental coordinates are reconstructed and coloured by the number of particles in clusters in a typical sample close to the cluster-gel line ($\phi=7.5~\%$, $c_p=1$ g/$\ell$). Origin of time is the last frame before melting of Wigner crystal.
	%Phase diagram obtained in reservoir cell. Spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of \textbf{b-c} is highlighted by a square.
	}
	\label{fig:cell_vs_cap}
\end{figure*}

\clearpage 
\begin{figure}
\includegraphics[width=10cm]{figure2.pdf}
\caption{\textbf{Different regimes of gelation.} \textbf{a,} Phase diagram. Experimental points are categorised based on the final state obtained in reservoir cell. The spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of Fig.~\ref{fig:cell_vs_cap}b-c is highlighted by a square. \textbf{b,} Comparison of system evolution in terms of largest cluster extent and of mean coordination number. By increasing density: $\phi=4.2,\,8,\,16,\,27~\%$, $c_p=1,\,1.5,\,1.2,\, 1$ g/$\ell$ for 
%\textcolor{red!40!yellow}
{$\blacklozenge$}, 
%\textcolor{red!80!yellow}
{$\circ$}, 
%\textcolor{red!80!black}
{\tiny$\blacksquare$} and 
%\textcolor{red!40!black}
{$\bullet$} respectively. \textbf{c,} Growth of the characteristic wave number for the same samples.}
\label{fig:phasediag}
\end{figure}

\clearpage
\begin{figure}
	\includegraphics[width=10cm]{figure3.pdf}
	\caption{\textbf{Hydrodynamics} \textbf{a,} Probability of staying elongated for a triplet in a non-percolating sample ($\phi=4~\%$, $c_p=1$ g/$\ell$). The continuous line is the best exponential fit of characteristic time $27\tau_B$. Inset: probability distribution of radii of gyration of triplets. \textbf{b,} Evolution of the aspect ratios of clusters of 4 particles and more in the same sample (dashed lines) and in a percolating sample ($\phi=8~\%$, $c_p=1.5$ g/$\ell$, continuous lines). \textbf{c,} Bond angle distribution relative to existing bonds (gray), to a future bond (red) or to a future bond involving an isolated particle (blue) obtained in the percolating sample. Future bonds are shifted to smaller angles, whereas gas adsorption takes place from larger angles. Insets sketch both cases, with present bonds drawn thick and future bonds drawn dotted.}
	\label{fig:hydro}
\end{figure}

\clearpage
\begin{figure*}
\includegraphics[width=16cm]{figure4.pdf}
\caption{\textbf{Mechanical tension drives coarsening} 
\textbf{a,} Reconstruction from experimental coordinates ($\phi=29~\%$, $c_p=0.7$ g/$\ell$) of coarsening process geometry (top) and topology (bottom). Particles are drawn to scale and coloured by $q_2$ from blue (low) to red (high). The meshed surface is a Gaussian coarse-graining of the network pattern. The red line indicates the shortest on-graph path between the two particles of interest. 
\textbf{b,} Number of bond breaking function of the future on-graph distance $\Lambda$ (after percolation) for the three percolating state points of Fig. 2a (same colours and markers).
%three state points of increasing density: $\phi=8,\,16,\,27~\%$, $c_p=1.5,\,1.2,\,1$g/$\ell$ for \textcolor{red!80!yellow}{$\circ$}, \textcolor{red!80!black}{\tiny$\blacksquare$} and \textcolor{red!40!black}{$\bullet$} respectively.
\textbf{c,} Probability for a bond to break during $3\tau_B$ function of bond-centred $q_2$ for all events (blue) and for events with $\Lambda>1$ (thick red) in the same sample as \textbf{a}.
\textbf{d,} Evolution of the topologically relevant bond breaking probability ($\Lambda>3$) for the same state points as in b. Vertical lines indicate the respective percolation times.
}
\label{fig:breaking}
\end{figure*}

>>>>

The mechanism of spinodal gelation has particularly been studied for colloidal suspensions since they can be regarded as a model system for other systems. Thanks to extensive studies in the past, we now have a reasonable physical picture: Gelation is induced by spinodal decomposition into two phases whose dynamics are very different \cite{tanaka1999colloid,tanaka2000viscoelastic} and the resulting vitrification of the slower phase \cite{piazza1994phase,verhaegh1997,tanaka1999colloid,poon2002,lu2008gelation}. 
However, the knowledge of ordinary spinodal decomposition cannot explain the observed behaviour since it predicts that the minority colloid-rich phase should form isolated clusters rather than a percolated network. This mystery was solved by using an analogy of phase separation of colloidal suspensions to that of polymer solutions on the basis of the concept of dynamic asymmetry \cite{tanaka1999colloid}. 
What is common in these two systems is that a mixture is composed of large slow components (colloid or polymer) and small fast components (solvent). 
On a macroscopic level, such dynamically asymmetric mixtures are known to show universal pattern formation, known as ``viscoelastic phase separation'' \cite{tanaka2000viscoelastic}. 
On a microscopic level, however, there are essential differences between colloidal suspensions and polymer solutions coming from the topological difference (sphere vs. chain), leading to 
distinct rheological behaviours. Thus there are many open questions about the elementary process of phase separation of colloidal suspensions and the resulting colloidal gelation.   
From the physical viewpoint, the presence of a fluid component indicates the importance of fluid momentum conservation in phase separation, which makes hydrodynamics and mechanics active in structural evolution. This is the key for viscoelastic phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic}. There have been some numerical studies on the role of hydrodynamics~\cite{tanaka2000simulation,Furukawa2010} and mechanics~\cite{Tanaka2007,colombo2014self,testard2014intermittent} in colloidal gelation, however, there has so far been no experimental study that directly addresses these effects by dynamical observation of the process.

@@

To access roles of mechanics in the late stage, we realize a density matching of the order of 10$^{-4}$ (gravitational Peclet number $Pe< 10^{-7}$) and use a very accurate particle tracking method \cite{leocmach2013novel}. 

@@

The initial similarity between the trajectories of dilute gels and cluster phase is striking, as if their dynamics were ruled by the same phenomenon. To confront this hypothesis, we compute the time dependent static structure factor $S(q,t)$. For all gel and cluster samples, we observe the appearance of a low $q$ peak in $S(q)$, see Supplementary Figure XXX, which is characteristic of spinodal decomposition in a system of a conserved order parameter. To track the position of this peak, we compute the characteristic wave number $\langle q \rangle$ and show its temporal evolution in Fig.~\ref{fig:phasediag}c. The curves for all samples follow a master curve coherent with spinodal decomposition kinetics: At short times $\langle q \rangle(t)$ shows a plateau indicating that the low $q$ peak builds up at constant wave number corresponding to distances of about $2\sigma$. This plateau is characteristic of the early stage spinodal decomposition, which is described by Cahn's linear theory \cite{onuki2002phase}. At intermediate times, on the other hand, we observe coarsening with \textcolor{blue}{$\langle q \rangle \sim t^{-1/2}$, which is often observed in viscoelastic phase separation 
(see, e.g., \cite{Furukawa2010}).} Finally at longer times each sample deviates from the master curve to form a plateau indicating arrest. The more dilute samples arrest sooner, 
but the origin of arrest is different between clusters and percolated networks. For clusters, density fluctuations grow rather rapidly in the initial stage, but after the formation of isolated clusters their growth takes a long time since the diffusive evaporation-condensation mechanism does not work for colloidal viscoelastic phase separation and direct cluster collisions and coalescence are necessary for coarsening \cite{tanaka2000viscoelastic}. For networks, densification due to phase separation leads to dynamic arrest due to vitrification \cite{piazza1994phase,verhaegh1997,tanaka1999colloid,poon2002,zaccarelli2007,lu2008gelation} or formation of locally stable structures \cite{royall2008g}. 
Our observations indicate that both cluster and gel phases are due to viscoelastic spinodal decomposition: network-type spinodal for the gel, and droplet-type spinodal for the clusters where strong dynamical asymmetry between colloids and the solvent leads to unique roles of hydrodynamics and mechanics in phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic}.

@@

\clearpage
\begin{figure*}
	%\includegraphics[width=16cm]{figure1.pdf}
	\includegraphics{figs/cell_vs_cap2.pdf}
	\caption{\textbf{Reservoir cell.} \textbf{a,} Sketch of our experimental setup. The observation cell contains initially colloids, polymer and no salt. \textbf{b,} Confocal slice of a gel formed \textit{in situ} by our method ($\phi=25.5~\%$, $c_p=1.4$ g/$\ell$), 1 hour after gelation. \textbf{c,} Idem for a gel at the same state point formed \textit{ex situ} and immediately pumped into a capillary. \textbf{d,} Gelation observed by our method. Experimental coordinates are reconstructed and coloured by the number of particles in clusters in a typical sample close to the cluster-gel line ($\phi=7.5~\%$, $c_p=1$ g/$\ell$). Origin of time is the last frame before melting of Wigner crystal.
	%Phase diagram obtained in reservoir cell. Spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of \textbf{b-c} is highlighted by a square.
	}
	\label{fig:cell_vs_cap}
\end{figure*}

\clearpage 
\begin{figure}
%\includegraphics[width=10cm]{figure2.pdf}
\includegraphics{figs/phasediag.pdf}
\caption{\textbf{Different regimes of gelation.} \textbf{a,} Phase diagram. Experimental points are categorised based on the final state obtained in reservoir cell. The spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of Fig.~\ref{fig:cell_vs_cap}b-c is highlighted by a square. \textbf{b,} Comparison of system evolution in terms of largest cluster extent and of mean coordination number. By increasing density: $\phi=4.2,\,8,\,16,\,27~\%$, $c_p=1,\,1.5,\,1.2,\, 1$ g/$\ell$ for 
%\textcolor{red!40!yellow}
{$\blacklozenge$}, 
%\textcolor{red!80!yellow}
{$\circ$}, 
%\textcolor{red!80!black}
{\tiny$\blacksquare$} and 
%\textcolor{red!40!black}
{$\bullet$} respectively. \textbf{c,} Growth of the characteristic wave number for the same samples.}
\label{fig:phasediag}
\end{figure}

\clearpage
\begin{figure}
	%\includegraphics[width=10cm]{figure3.pdf}
	\includegraphics{figs/hydro.pdf}
	\caption{\textbf{Hydrodynamics} \textbf{a,} Probability of staying elongated for a triplet in a non-percolating sample ($\phi=4~\%$, $c_p=1$ g/$\ell$). The continuous line is the best exponential fit of characteristic time $27\tau_B$. Inset: probability distribution of radii of gyration of triplets. \textbf{b,} Evolution of the aspect ratios of clusters of 4 particles and more in the same sample (dashed lines) and in a percolating sample ($\phi=8~\%$, $c_p=1.5$ g/$\ell$, continuous lines). \textbf{c,} Bond angle distribution relative to existing bonds (gray), to a future bond (red) or to a future bond involving an isolated particle (blue) obtained in the percolating sample. Future bonds are shifted to smaller angles, whereas gas adsorption takes place from larger angles. Insets sketch both cases, with present bonds drawn thick and future bonds drawn dotted.}
	\label{fig:hydro}
\end{figure}

\clearpage
\begin{figure*}
%\includegraphics[width=16cm]{figure4.pdf}
\includegraphics{figs/breaking.pdf}
\caption{\textbf{Mechanical tension drives coarsening} 
\textbf{a,} Reconstruction from experimental coordinates ($\phi=29~\%$, $c_p=0.7$ g/$\ell$) of coarsening process geometry (top) and topology (bottom). Particles are drawn to scale and coloured by $q_2$ from blue (low) to red (high). The meshed surface is a Gaussian coarse-graining of the network pattern. The red line indicates the shortest on-graph path between the two particles of interest. 
\textbf{b,} Number of bond breaking function of the future on-graph distance $\Lambda$ (after percolation) for the three percolating state points of Fig. 2a (same colours and markers).
%three state points of increasing density: $\phi=8,\,16,\,27~\%$, $c_p=1.5,\,1.2,\,1$g/$\ell$ for \textcolor{red!80!yellow}{$\circ$}, \textcolor{red!80!black}{\tiny$\blacksquare$} and \textcolor{red!40!black}{$\bullet$} respectively.
\textbf{c,} Probability for a bond to break during $3\tau_B$ function of bond-centred $q_2$ for all events (blue) and for events with $\Lambda>1$ (thick red) in the same sample as \textbf{a}.
\textbf{d,} Evolution of the topologically relevant bond breaking probability ($\Lambda>3$) for the same state points as in b. Vertical lines indicate the respective percolation times.
}
\label{fig:breaking}
\end{figure*}
